---
title: 'Capstone project 2: Choose Your Own'
author: "Alicia Alfonso"
date: "2022-11-20"
output:
  pdf_document: default
  html_document: default
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

# I. Introduction

Daphniids are small crustaceans that live in freshwater ecosystems. Like
many aquatic invertebrates, their body is composed of calcium (Ca) and
their survival likely depends on the availability of this element in the
water. The dataset used for this project are simplified results from a
personal experiment (Alfonso, A. 2018 unpublished). The aim of the
experiment was to assess the inter-clonal variation between two
populations/clones (Northern and Southern) at different Ca
concentrations (Ca 0.5 mg/l, 4 mg/l and 106 mg/l) and temperatures (15
and 25ºC). With this information, we will create a model to predict the
survival of Daphniids evaluating the effect of the different variables
measured.

# II. Methods and analysis

## Part 1: Data exploration

```{r, error = FALSE}
data <- read.table("CapstoneProjectData.txt", header=TRUE) # Load dataset
library(survival) # For modelling
library(survminer) # For survival curves
library(dplyr) # For analysis
library(ggplot2) # For visualization

str(data) # 123 obs. of  6 variables
head(data, 10)
summary(data)


p <- data %>%
  ggplot(aes(x=Time)) +
    geom_histogram( binwidth=3, fill="#69b3a2", color="#e9ecef", alpha=0.9) +
    ggtitle("Final age") +
    theme(plot.title = element_text(size=15))
p

sum(data$Time == 88) # 13 alive at end

data %>%
  group_by(Clone, Temp, Ca) %>%
  summarise_at(vars(Time), list(name = mean))
```

The table is composed by 123 observations of 6 different variables:
Clone (Northern, Southern), Temperature (15ºC, 25ºC), Ca concentration
(0.5, 4, 106 mg/l), unique ID of the animal, Time (age at death or
maximum survival) and Death (TRUE, FALSE). The name of the clones refer
to the geographical origin of the animals, the temperature to the
degrees they were grown at (Celsius) and the Ca concentration the
different amounts of calcium present in the water. This dataset includes
123 animals followed through a maximum of 88 days. Time and Death then,
provide information on the day of their passing or the end of the
experiment. By the end of the experiment only 13 animals were alive and
the mean survival was 39 days.

Survival time was lowest at Ca 0.5, peaking at Ca 4, and declining again
at the highest Ca concentration. They also survived longer at 15C in
comparison to 25C. This indicates strong effects of Ca and temperature
that need to be taken into account for our model.

## Part 2: Modelling survival

For all variables, I started models with all factors independently and
all pairwise interactions (clone, temperature and Ca) to asses which
ones contributed the most. This analysis was done with the packages
survival (Therneau, 2015) and survminer (Kassambara & Kosinski, 2018)
for data fitting and representation. A stepwise model selection
discarded Ca × Temperature interaction, giving a more fitted
non-parametric Cox proportional hazard model with this form:

**log(hazard rate) \~ Clone + Temp + Ca + Clone × Temp + Clone × Ca**

The hazard rate is the probability of an animal dying within a given
time interval, given that it was alive at the start of it. As such, it
corresponds to the instantaneous mortality rate in a continuous-time
population model. The full model was contrasted against three other
simplifications of it, from which we chose the one with the lowest
Akaike Information Criterion (AIC). The final model is calculated with
95% confidence intervals of treatment effects.

```{r, error = FALSE}
s <- Surv(data$Time, data$Dead) # Creating our hazard rate
# Survival Plots
m1 <- survfit(s ~ 1, data = data, conf.type = "log-log") 
m1p <- ggsurvplot(m1, # Our base curve for all treatments
                  data = data,
                  palette = "#2E9FDF",
                  title = "Total survival",
                  size = .5,
                  ggtheme = theme_bw())

m2 <- survfit(s ~ Clone, data = data, conf.type = "log-log")
m2p <- ggsurvplot(m2, # Our curve per clonal population
                  data = data,
                  conf.int=TRUE,
                  title = "Survival per Clone",
                  size = .5,  # change line size
                  ggtheme = theme_bw(), # Change ggplot2 theme
                  palette = c("#A569BD", "#48C9B0"), # custom color palette
                  pval = TRUE)

m3 <- survfit(s ~ Temp, data = data, conf.type = "log-log")
m3p <- ggsurvplot(m3, # Our curve by temperature
                  data = data,
                  conf.int=TRUE,
                  title = "Survival per Temperature",
                  size = .5,  # change line size
                  ggtheme = theme_bw(), # Change ggplot2 theme
                  palette = c("#006399", "#CC0000"), # custom color palette
                  pval = TRUE)

m4 <- survfit(s ~ Ca, data = data, conf.type = "log-log")
m4p <- ggsurvplot(m4, # Our curve per Ca concentration
                  data = data,
                  conf.int=TRUE,
                  title = "Survival per Ca concentration",
                  size = .5,  # change line size
                  ggtheme = theme_bw(), # Change ggplot2 theme
                  palette = c("#F39C12","#E67E22","#D35400"),
                  pval = TRUE)

survplots <- list(m1p, m2p, m3p, m4p)
arrange_ggsurvplots(survplots, # Arrange multiple survplots
                    ncol = 2, nrow = 2)
# Median survival time is 39 days
# Overlapping between clones
# Overlapping between temperatures
# No-overlapping for 0.5 mg Ca/l
```

# IV. Results

A Cox regression model was fitted using initially all pairwise
interactions and chosen after backward selection using clone,
temperature and Ca as explanatory variables. For the different Cox
regression models tested, "Model 4" with lowest AIC value (769.51) was
accounted for being the best fitted. Consistently with our exploration,
and according to Hazard ratios, mortality probabilities are much higher
for extreme Ca concentrations (0.5 and 106), 25ºC, and the Northern
clone.

```{r, error = FALSE}
data$Clone <- as.factor(data$Clone)
data$Temp <- as.factor(data$Temp)
data$Ca <- as.factor(data$Ca)
# Non-parametric Cox proportional hazards models
# Fits a Cox proportional hazards regression model
summary(mod1 <- coxph(s ~ Clone + Temp * Ca, data=data)) # Ca:Temp interaction
summary(mod2 <- coxph(s ~ Clone + Temp + Ca, data=data)) # No interaction
summary(mod3 <- coxph(s ~ (Clone + Temp + Ca)^2, data=data)) # Ca:Temp, 
# Clone:Temp, Clone:Ca interaction

anova(mod1, mod2, mod3) # (!) Model 3 is already significant
AIC(mod1, mod2, mod3)   # 

# Using the step method to refine our best model (mod3) into a better one (mod4)
summary(mod4 <- step(mod3))
# Lowest AIC: Clone:Temp & Clone:Ca retained, no Temp:Ca interaction
AIC(mod1, mod2, mod3, mod4)
# Creating the better fitted model adding interactions Clone x Temp and 
# Clone x Ca
mod4 <- coxph(s ~ Clone + Temp + Ca + Clone:Temp + Clone:Ca, data=data)
summary(mod4)



# Model 4 visualization separated by treatment 
mod4S15 <- survfit(mod4, newdata=data.frame(Clone="Southern", Temp="15", 
                                            Ca=c("0.5", "4", "106")))
pmod4S15 <- ggsurvplot(mod4S15, conf.int=TRUE,
                     title = "Survival Southern 15C",
                     legend.labs = c("0.5", "4", "106"),
                     data=data.frame(Clone="Southern", Temp="15", 
                                     Ca=c("0.5", "4", "106")),
                     size = .5,  # change line size
                     xlab = "Time (days)",
                     ggtheme = theme_classic(), # Change ggplot2 theme
                     palette = c("#F39C12","#E67E22","#D35400"))

mod4S25 <- survfit(mod4, newdata=data.frame(Clone="Southern",Temp="25", 
                                            Ca=c("0.5", "4", "106")))
pmod4S25 <- ggsurvplot(mod4S25, conf.int=TRUE,
                     title = "Survival Southern 25C",
                     legend.labs = c("0.5", "4", "106"),
                     data=data.frame(Clone="Southern", Temp="25", 
                                     Ca=c("0.5", "4", "106")),
                     size = .5,  # change line size
                    xlab = "Time (days)",
                    ggtheme = theme_classic(), # Change ggplot2 theme
                     palette = c("#F39C12","#E67E22","#D35400"))

mod4N15 <- survfit(mod4, newdata=data.frame(Clone="Northern",Temp="15", 
                                            Ca=c("0.5", "4", "106")))
pmod4N15 <- ggsurvplot(mod4N15, conf.int=TRUE,
                     title = "Survival Northern 15C",
                     legend.labs = c("0.5", "4", "106"),
                     data=data.frame(Clone="Northern", Temp="15", 
                                     Ca=c("0.5", "4", "106")),
                     size = .5,  # change line size
                    xlab = "Time (days)",
                    ggtheme = theme_classic(), # Change ggplot2 theme
                     palette = c("#F39C12","#E67E22","#D35400"))

mod4N25 <- survfit(mod4, newdata=data.frame(Clone="Northern",Temp="25", 
                                            Ca=c("0.5", "4", "106")))
pmod4N25 <- ggsurvplot(mod4N25, conf.int=TRUE,
                     title = "Survival Northern 25C",
                     legend.labs = c("0.5", "4", "106"),
                     data=data.frame(Clone="Northern", Temp="25", 
                                     Ca=c("0.5", "4", "106")),
                     size = .5,  # change line size
                    xlab = "Time (days)",
                    ggtheme = theme_classic(), # Change ggplot2 theme
                     palette = c("#F39C12","#E67E22","#D35400"))

mod4plots <- list(pmod4S15, pmod4S25, pmod4N15, pmod4N25)
arrange_ggsurvplots(mod4plots, # Arrange multiple survplots
                    title = "Model 4",
                    ncol = 2, nrow = 2)
```

Survival curves for each Ca treatment (0.5, 4 and 106 mg Ca l-1),
temperature (15 and 25ºC) and clone population (Northern and Southern).
Shaded error bands represent 95% confidence intervals around the fitted
model (solid lines).

# V. Conclusion

Modelling survival of organisms can give us a deep insight into which
factors contribute to their success in an ecosystem. Using R for this
purpose proves extremely useful and practical as we are able to generate
predictions from smaller datasets. In our case, we establish the strong
negative effect that low calcium concentrations have in the survival of
Daphniids. We can conclude that the mortality will highly increase if
there is a depletion of this element and that northern populations will
be the most affected. Our model also hints towards strong interactions
between variables that leads to more interesting questions regarding
phenotypic plasticity or even toxicity for even higher calcium
concentrations.
